The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Tides or nature?

### 2025-08-29 thesis/abstract.tex
julio edits at durham

Dwarf galaxies are unique cosmological probes, testing our understanding
of the smallest scales of structure formation in the Universe. Among the
classical dwarf galaxies, Sculptor and Ursa Minor appear unusual, with
more stars than expected in the outer regions compared with the typical
Exponential density profile, rememnescent of tidal interaction. I verify
the validity of this excess, checking \citet{jensen+2024}'s bayesian
membership catalogue for completeness, simple cuts, and other surveys.
To evaluate if tides indeed can explain these density features, I
conduct idealized N-body simulation of both galaxies in the tidal field
of the Milky Way. In both cases, the orbits are well constrained,
producing pericenters larger than expected to produce substantial tidal
impacts. In addition, the time since the last pericenter does not agree
with a tidal interpretation. However, Sculptor's orbit is strongly
influenced by the presence of an LMC. With an LMC, Sculptor's long term
orbital history is highly uncertain, however recent tidal effects due to
the Milky Way and LMC only minimally affect this galaxy. Because tides
likely do not explain the presence of these density features, I review
other theories about the formation of dwarf galaxy ``halos,'' including
mergers / accretion events and complex multi-component star formation
histories. The density profiles of Sculptor and Ursa Minor, with their
deviations from the classically expected exponential, present an
interesting puzzle, likely with a solution involving their formation and
history in the early universe.

>>>>

The satellite galaxies of the Milky Way (MW), with the exception of the
Magellanic clouds, are all dwarf spheroidals (dSphs)---gas-free,
non-rotating stellar systems who's gravity is dominated by dark matter.
The light profiles of dSphs are usually well-approximated by an
exponential law, with a sharp outer cutoff. Yet systems like the
Sculptor (Scl) and Ursa Minor (UMi) dSphs have been found to host member
stars far from the main body, out to \textasciitilde10 effective radii.
The origin of these extended profiles is not clear, and may reflect
either the influence of Galactic tides or instead an innate feature. In
this thesis, I review the excestence of an extended light profiles in
Scl and UMi, validating the Bayesian membership catalogue of
\citet{jensen+2024}. To evaluate if tides indeed can explain these
density features, I conduct idealized N-body simulation of both galaxies
in the tidal field of the Milky Way. I find that neigher dwarf has been
subjected to tidal forces strong enough to create the observed extended
light profiles from initially exponential profiles. One complication is
that Sculptor's orbit is strongly influenced by the presence of the
Large Magellanic Cloud (LMC). Because the LMC's mass is poorly
constrained, Sculptor's long term orbital history is highly uncertain.
Despite this uncertainty, our models still suggest that the combined
tides of the LMC and MW are still unable to explain Scl's outer profile
over the past 5 Gyr. I conclude that the extended light profiles of Scl
and UMi are innate, likely reflecting their formation histories.
Mergers, accretion events, or complex, multi-component star formation
episodes are possible explanations for such an extended outer profile.

### 2025-08-29 thesis/data.tex
julio edits at durham

To provide an empirical benchmark for simulations in the next chapters,
we aim to understand the observed properties and density profiles of
Sculptor and Ursa Minor. In this section, we first compile past
measurements of each galaxy. We then describe the \citet{jensen+2024}
Bayesian methodology to select high probability member stars from
\emph{Gaia}. We show the derived density profiles are robust to changes
in assumptions. While Fornax is well represented by a 2D exponential
profile, Sculptor and Ursa Minor show an excess of stars relative to an
2D exponential in the outer regions, better described by a Plummer
profile. Future chapters assess possible explanations for these excess.

\section{\texorpdfstring{\emph{Gaia} data}{Gaia data}}\label{gaia-data}

Here, we briefly describe \citet{jensen+2024}'s (hereafter J+24)
membership estimation method. J+24 use a Bayesian framework
incorporating proper motion (PM), colour-magnitude diagram (CMD), and
spatial information to determine the probability that a given star
belongs to the satellite. By accounting for PM in particular, J+24
produces low contamination samples of candidate member stars out to
large distances from a dwarf galaxy. J+24 extends the algorithm
presented in \citet{MV2020a}; \citet{MV2020b} but additionally includes
an optional secondary, extended spatial component to find possible
members as far as \textasciitilde10 half-light radii \(R_h\) from
several dwarf galaxies. See also similar work by \citet{pace+li2019};
\citet{battaglia+2022}; \citet{pace+erkal+li2022}; \citet{qi+2022}.

J+24 define likelihoods \({\cal L}\) representing the probability
density that a star is consistent with either the MW stellar background
(\({\cal L}_{\rm bg}\)) or the satellite galaxy
(\({\cal L}_{\rm sat}\)). In either case, the likelihoods are the
product of a spatial, PM, and CMD term: \begin{equation}{
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
}\end{equation}

Each likelihood is normalized over their respective 2D parameter space
for both the satellite. To control the relative frequency of member and
background stars, \(f_{\rm sat}\) representing the fraction of member
stars in the field. The total likelihood for any star in this model is
the sum of the satellite and background likelihoods, weighted by their
relative frequencies, \begin{equation}{
{\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}.
}\end{equation} The probability that any star belongs to the satellite
is then given by \begin{equation}{
P_{\rm sat} = 
\frac{f_{\rm sat}\,{\cal L}_{\rm sat}}{{\cal L}_{\rm tot}}
= \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}.
}\end{equation}

\section{Selected stars}\label{selected-stars}

In
Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection},
we illustrate the resulting samples from the algorithm in the tangent
plane, CMD, and proper motion space. The tangent plane coordinates
\(\xi\) and \(\eta\) are the distances in RA and declination as measured
on a plane tangent to the dwarf galaxies centre. \(R_{\rm ell}\) is
defined as the circularized elliptical radius, so
\(R_{\rm ell}^2 = a\,b\,({\xi'}^2 / a^2 + {\eta'}^2 / b^2)\) where
\(\xi'\) and \(\eta'\) are rotated to align with the position angle
direction \(\theta\).

For our fiducial sample, we adopt a probability cut of
\(P_{\rm sat} = 0.2\). Most stars are assigned probabilities close to
either 0 or 1, so the choice of probability threshold is only marginally
significant. Additionally, even for a probability cut of 0.2, the purity
of the resulting sample with RV measurements is very high
(\textasciitilde90\%, J+24). Note there is likely a systematic bias in
using stars with RV measurements to measure purity. Fainter stars are
less likely to have been targeted and have poorer astrometry and colour
information. We find no difference in the resulting density
distributions when restricting stars to be brighter than a specific
magnitude.

For all galaxies, each criteria is commensurate in sifting out
nonmembers. The CMD is well defined, and probable members only extend a
few times the colour uncertainty from the CMD. In proper motion space,
selected stars are within \(\masyr\) from the systematic proper motion.
The stars furthest away in proper motion space typically have large
uncertainties \(\sim 1\masyr\), so most members are reasonably
consistent. In any case, the algorithm removes stars with inconsistent
PMs. Finally, the spatial likelihood reduces the probabilities of stars
distant from the dwarfs centres. Member stars are rare outside of
\(\sim5R_h\), resulting from both the likelihood specification and the
lower density of background stars consistent with the CMD and PM of the
satellite.

One limitation of the J+24 method is that it assumes a spatial form of
the dwarf galaxy, possibly limiting the detection of very distant stars
or features. Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection} also
illustrates the distribution of stars selected without a spatial
criterion. We define the \emph{CMD+PM} selection as
\begin{equation}\protect\phantomsection\label{eq:sel_cmd_pm}{
{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}
}\end{equation} with the likelihoods from J+24 as described above. These
stars are distributed similar to the fiducial (probable members) sample
but with a background uniform distribution across the entire field. This
illustrates the approximate background of stars which may be confused as
members. Additionally, since there is no clear spatial structure in the
\emph{CMD+PM} (or \emph{all}) sample outside several \(R_h\), it is
unlikely that there are additional faint, tidal features detectable with
\emph{Gaia}. Not shown here, we also try a variety of simpler, absolute
cuts and thresholds, finding no evidence of other structure in both
\emph{Gaia} and DELVE/UNIONS data.

Finally, we illustrate the location of RV-confirmed members from
Section~\ref{sec:rv_obs} in each panel. Because RV targets tend to be
brighter than a typical \emph{Gaia} candidate, RV members typically have
more precise PMs. However, the RV members fill about the same area on
the CMD down to the magnitude limit. Additionally, while the number of
stars with RV measurements decreases at large radii, there are still
confirmed RV members as far as \(\sim 10R_h\) for both galaxies. These
stars illustrate the extended profiles of each galaxy. Note that for an
exponential profile, 99.95\% of stars should be within \(6R_h\), so the
discovery of any stars beyond \(6R_h\) hints at a deviation from an
exponential profile. These \emph{extratidal} stars hint at something
tidal in nature, and our goal is to explore possible interpretations.

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/scl_selection.png}
\caption[Sculptor sample selection]{The selection criteria for Scl
members. Light grey points represent field stars (satisfying quality
criteria), turquoise points CMD+PM selected stars
(Eq.~\ref{eq:sel_cmd_pm}), blue xs probable members (2-component), and
RV confirmed members indigo diamonds. We mark the two stars from
\citet{sestito+2023a} with rust-outlined indigo stars. \textbf{Top:}
Tangent plane \(\xi, \eta\). The orange ellipse represents 3 half-light
radii. \textbf{Bottom left:} Colour magnitude diagram in Gaia \(G\)
magnitude versus \(G_{\rm BP} - G_{\rm RP}\) colour. We plot the Padova
12Gyr {[}Fe/H{]}=-1.68 isochone in orange. The black bar in the top left
represents the median colour error. \textbf{Bottom right:} Proper motion
in declination \(\mu_\delta\) vs RA \(\mu_{\alpha*}\) (corrected) the
orange circle represents the \citet{MV2020b} proper motion. The black
cross represents the median proper motion
error.}\label{fig:scl_selection}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/umi_selection.png}
\caption[Ursa Minor sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Ursa Minor. We outline RV
members outside of \(6R_h\) in black stars (from \citet{sestito+2023b},
\citet{pace+2020} and \citet{spencer+2018}).}\label{fig:umi_selection}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/fornax_selection.png}
\caption[Fornax sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Fornax. RV members are from
\citet{WMO2009}. While possibly a limitation of RV sample selection,
Fornax does not show the same extended outer halo of probable members as
Sculptor or Ursa Minor despite having many more
stars.}\label{fig:fornax_selection}
\end{figure}

@@

Density profiles are an essential observational constraint for our later
simulations. To derive density profiles, we use 0.05 dex bins in log
radius. We remove bins at smaller (larger) radii than the first empty
bin working outwards. We use symmetric poisson uncertainties. As
discussed below, these uncertainties are straightforward but are likely
under-representative. We use the values from \citet{munoz+2018}'s Sérsic
maximum likelihood fits for \(R_h\), which are more precisely derived
given their deeper photometry.

Fig.~\ref{fig:scl_observed_profiles} show the derived density profiles
for Sculptor, Ursa Minor, and Fornax. We calculate density profiles for
different selections of stars from above: all quality stars, CMD+PM
only, and the probable member samples. In each case, all samples are the
same towards the inner regions of the satellite. Classical dwarfs
dominate the stellar density in their cores. However, the density
profile \emph{all stars} plateaus at the total background in the field
at radii of 30-60 arcminutes. By restricting stars to being most likely
satellite members by CMD + PM, the background is reduced by 1-2 dex. The
CMD+PM background plateau likely represents the density of background
stars which could be mistaken as members. Finally, the background
(BG)-subtracted profile results from subtracting the apparent background
in the \emph{all} profile.

Note that the probable members (fiducial) density profile continues to
confidently estimate the density profile below the CMD+PM background.
These points are likely unreliable (see discussion below). However,
before this point, both the BG subtracted and probable members density
profiles are strikingly similar. Assumptions about the details of the
likelihood and spatial dependence have marginal influence on the
resulting density profile when the satellite is higher density than the
background. Thus the detection of excesses of stars in Sculptor (past 20
arcmin) and UMi (past 15-20 arcmin) are robust.

Nearby to UMi, there is a small, likely unassociated, ultrafaint star
cluster, Muñoz 1 \citep{munoz+2012}. The cluster is at a relative
position of \((\xi, \eta) \approx(-42, -15)\) arcminutes, corresponding
to an elliptical radius of 36 arcminutes. However this cluster does not
have a bright RGB, so has few stars brighter than a \(G\) mag of 22.
While Muñoz 1 may contribute \(\sim\) 5 stars to the density profile of
UMi, this would possibly only affect 1-2 bins due to its compact size.

To illustrate the differences between each dwarf galaxy, in
Fig.~\ref{fig:classical_dwarfs_densities}, we compare Sculptor, Ursa
Minor, and other classical dwarfs against 2D-exponential and Plummer
density profiles (REF). All dwarf galaxies appear to be well described
by an exponential profile in the inner regions. However, while an
exponential continues to be a good description for other dwarf galaxies,
Sculptor and Ursa Minor diverge, showing a substantial excess over an
exponential, better fit by a Plummer instead. In fact, Sculptor and Ursa
Minor show excesses of between 2-3 orders of magnitude beyond what is
expected of an exponential at the very outer regions!

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_fnx_density_methods.pdf}}
\caption[Sculptor density profiles]{The density profile of Sculptor,
Ursa Minor, and Fornax for different selection criteria, ploted as log
surface dencity versus log elliptical radius. \emph{CMD+PM} select stars
more likely to be members according to CMD and PM only, \emph{all}
selects any high quality star, and \emph{BG subtracted} is the
background-subtracted density derived from \emph{all} stars. We mark the
half-light radius (vertical dashed line) and the break radius (black
arrow, REF).}\label{fig:scl_observed_profiles}
\end{figure}

@@

\section{Summary}\label{summary}

In summary, we have used J+24 data to derive the density profiles for
the classical dwarf galaxies. In each case, the density profile is
robust against different selection criteria. Both Sculptor and Ursa
Minor show strong evidence for deviations from an exponential profile.
We find no evidence of additional (velocity or stellar) substructure in
either galaxy. Our density profiles are consistent with many previous
works, all indicating divergence from a King or Exponential density
profile with a break between 30 and 60 arcminutes. Our goal in the
following chapters is to test if tides provide a viable explanation for
the observed properties of Sculptor and Ursa Minor.

>>>>

As discussed in Section~\ref{sec:exponential_profiles}, the projected
luminosity/stellar mass density profile of dwarf galaxies is generally
well-described by an exponential law Eq.~\ref{eq:exponential_profiles}.
A prototypical example is the Fornax dSph --- well-fit by an exponential
over 4.5 decades in surface density. On the other hand, dSph like
Sculptor or Ursa Minor have profiles with deviate significantly from an
exponential profile fitted to the inner regions, with a clear excess of
stars/light in the outer regions. In this chapter, we critically review
the density profile of MW classical dwarfs. This thesis concerns the
origin of this excess, with a tidal interpretation explored in
subsequent chapters.

\textbf{question: how much about density profiles in this
paragraph\ldots{}}

\section{\texorpdfstring{Satellite stellar membership with
\emph{Gaia}}{Satellite stellar membership with Gaia}}\label{satellite-stellar-membership-with-gaia}

Measuring the light profile of a resolved galaxy requires careful
consideration of whether a given star belongs to the system. Without
removing contamination from foreground/background features, faint
features may be lost in the noise or be of uncertain association. When
only photometric data was available, the galaxy membership of stars were
ascertained using the colour-magnitude diagram alone \citep[e.g.,
matched filter methods like][]{rockosi+2002}. Now that \emph{Gaia} data
are available, stellar parallax and proper motion are also available to
improve membership assignment.

Here, we use the \citet{jensen+2024}'s (hereafter J+24) membership
probabilities for \emph{Gaia} observations. J+24 used a Bayesian
framework incorporating proper motion (PM), colour-magnitude diagram
(CMD), and spatial information to determine the probability that a given
star belongs to the satellite or foreground/background. By accounting
for PM in particular, J+24 produced low contamination samples of
candidate member stars out to large distances from a dwarf galaxy. J+24
extended the algorithm presented in \citet{MV2020a}; \citet{MV2020b} by
additionally includes a secondary, extended spatial component. J+24
detected candidate members out to \textasciitilde10 half-light radii
from the centres of some galaxies (\(R_h\)). Similar recent work has
also included \citet{pace+li2019}; \citet{battaglia+2022};
\citet{pace+erkal+li2022}; \citet{qi+2022}.

J+24 construct likelihoods, \({\cal L}\), representing the probability
density that a star is consistent with either the foreground/background,
\({\cal L}_{\rm bg}\), or the satellite galaxy, \({\cal L}_{\rm sat}\).
In either case, the likelihoods are the product of a spatial, PM, and
CMD component, \begin{equation}{
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
}\end{equation}

For the satellite, the \emph{spatial likelihood} is specified either as
a one- or two-component elliptical exponential profile. A second
component is included only when the preferred amplitude of the second
component is non-zero. The \emph{proper-motion likelihood} quantifies
the agreement of a star's motion with the dwarf galaxy's systemic
motion, accounting for observational uncertainties. The \emph{CMD
likelihood} measures the consistency of a star's \emph{Gaia} \(G\)
magnitude and \(G_{\rm BP}- G_{\rm RP}\) colour with theoretical
isochrones for the galaxy. For the background likelihoods, the spatial
likelihood is a uniform distribution over the field, and the background
CMD and PM likelihoods are constructed empirically from the stars in an
annulus far from the satellite. Each likelihood is normalized as a
probability density over the respective parameter space.

The total likelihood is a mixture model of the satellite and background,
weighted by the fraction of stars in the field belonging to the
satellite, \(f_{\rm sat}\): \begin{equation}{
{\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}.
}\end{equation} The probability that a given star belongs to the
satellite is then \begin{equation}{
P_{\rm sat} = 
\frac{f_{\rm sat}\,{\cal L}_{\rm sat}}{{\cal L}_{\rm tot}}
= \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}.
}\end{equation}

J+24 fit this model using Monte Carlo Markov chain simulations, solving
for the proper motions, satellite membership fraction (\(f_{\rm sat}\)),
and structural properties of the second exponential density profile (if
included). The median parameters from the samples are then used to
calculate the \(P_{\rm sat}\) we use for sample selection.

For our fiducial sample, we adopt a minimum probability of
\(P_{\rm sat} = 0.2\). Most stars have \(P_{\rm  sat}\) values which are
nearly 0 or 1, so the exact choice of probability threshold has little
effect on the resulting sample. Even at our relatively generous
probability threshold of 0.2, the purity remains high when validated
against spectroscopic line-of-sight (LOS) velocities
(\textasciitilde90\%, J+24). We note, however, that these purity
estimates may be biased. Stars with LOS velocities are typically
brighter and, as a result, have more precise \emph{Gaia} measurements.
However, we find that our conclusions are unchanged when limiting
samples to only the brightest stars. Altogether, the J+24 method
provides a high-quality, low-contamination sample of dwarf galaxy
candidate member stars, which we will now investigate.

\section{The effects of membership
criteria}\label{the-effects-of-membership-criteria}

We analyze stellar distributions in the tangent plane, considering as
well the projected shape of a galaxy. The tangent plane coordinates
\(\xi\) and \(\eta\) are offsets in RA and declination as measured on
the plane tangent to the galaxy centre. To account for the elliptical
shape of the galaxy, we use \(R_{\rm ell}\), which we define as the
circularized elliptical radius, \begin{equation}{
R_{\rm ell}^2 = a\,b\,({\xi'}^2 / a^2 + {\eta'}^2 / b^2),
}\end{equation} where \(\xi'\) and \(\eta'\) are the tangent plane
coordinate rotated to align with a dwarf galaxy's major and minor axis,
and \(a\) and \(b\) are the semi-major and semi-minor axis of the
galaxy.

We illustrate the likelihood-based membership selection criteria through
a progressive tightening of criteria. First, we consider a
minimally-refined sample, \textbf{all}, only excluding stars with poor
astrometry, unreliable photometry, or inconsistent parallaxes. Next,
incorporating CMD and PM information, we use the \textbf{CMD+PM} to test
a selection method agnostic to the spatial position. This sample
includes stars where the CMD and PM combined likelihood favours
satellite membership:
\begin{equation}\protect\phantomsection\label{eq:sel_cmd_pm}{
{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}.
}\end{equation} Next, our \textbf{fiducial sample},
\(P_{\rm sat} > 0.2\), includes a spatial likelihood as described in
J+24. Finally, the \textbf{RV members} sample adds line-of-sight
velocity information (see Appendix for details). While the latter is the
best motivated membership sample, we do not use this sample for stellar
density analysis due to its incompleteness and complex selection
function.

Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}
shows our fiducial sample for each galaxy, and the impact of alternate
samples, in tangent-plane coordinates, the CMD, and in proper-motion
space. The \emph{all} sample extends uniformly across the tangent plane,
but includes a substantial background population. The \emph{CMD+PM}
sample has a much lower background density in the tangent plane,
revealing the satellite more clearly. The fiducial sample appears
similar to the \emph{CMD+PM} sample in the CMD and PM planes, but
excludes improbably distant stars. The \emph{RV member} sample also
traces out similar distributions in CMD and PM space as the fiducial
sample, with less dispersion, likely reflecting the brighter magnitudes
of stars with spectroscopic follow-up. Each component plays an important
role in ensuring a high-quality membership criterion.

Based on the fiducial sample's distribution in
Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection},
Ursa Minor stands out because of its higher ellipticity. Sculptor and
Ursa Minor show a population of members past \(6R_h\) in radius, but
Fornax does not, even though Fornax has many more stars. Fornax also has
a bluer CMD, indicative of more recent star formation. Otherwise, there
are no standout morphological features between these galaxies.

One limitation of the J+24 method is the assumption of a specific
density profile for the dwarf galaxy (a one- or two-component
exponential), which may impact the membership probability of distant
stars. However, there are no clear extensions or overdensities in the
\emph{CMD+PM} selected sample, which does not include a spatial
likelihood. If even fainter, more extended features exist, they are
apparently not detectable with \emph{Gaia}.

To illustrate the extent of Sculptor and Ursa Minor, we highlight the
most distant, spectroscopically confirmed members with red-outlined
stars in Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}. In
particular, \citet{sestito+2023a}; \citet{sestito+2023b} targeted the
most distant stars using the J+24 candidate membership list. These stars
illustrate an extended stellar distribution for each galaxy. For an
exponential profile, 99.95\% of stars fall within a radius of \(6R_h\),
so the mere presence of very distant member stars provides direct
evidence for deviations from a simple, exponential profile.

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/scl_selection.png}
\caption[Sculptor sample selection]{The distributions of various samples
of \emph{Gaia} stars for Sculptor. We plot light grey points for field
stars (with consistent parallaxes and reliable photometry and
astrometry), turquoise points for \emph{CMD+PM} selected stars
(Eq.~\ref{eq:sel_cmd_pm}), blue \texttt{x}s for the fiducial sample
(\(P_{\rm sat} > 0.2\)), and indigo diamonds for the RV confirmed
members. We mark the two far-outlier stars from \citet{sestito+2023a}
with rust-outlined indigo stars. \textbf{Top:} Tangent plane
\(\xi, \eta\). The orange ellipse represents 3 half-light radii.
\textbf{Bottom left:} Colour magnitude diagram in Gaia \(G\) magnitude
versus \(G_{\rm BP} - G_{\rm RP}\) colour. We plot a Padova \(12\,\)Gyr,
\({\rm [Fe/H] }=-1.68\) isochone in orange. The black bar in the top
left represents the median colour error. \textbf{Bottom right:} Proper
motion in declination \(\mu_\delta\) vs RA \(\mu_{\alpha*}\)
(corrected). The orange point marks the systemic \citet{MV2020b} proper
motion. The black cross represents the median proper motion error.
\textbf{TODO: reduce png sizes}}\label{fig:scl_selection}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/umi_selection.png}
\caption[Ursa Minor sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Ursa Minor. We outline RV
members outside a radius of \(6R_h\) with red stars (from
\citet{sestito+2023b}, \citet{pace+2020} and \citet{spencer+2018}).
\textbf{TODO: plot 6Rh ellipse\ldots{}}}\label{fig:umi_selection}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/fornax_selection.png}
\caption[Fornax sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Fornax. RV members are from
\citet{WMO2009}. Fornax does not show the same extended outer halo of
probable members as Sculptor or Ursa Minor despite having many more
stars.}\label{fig:fornax_selection}
\end{figure}

@@

We derive density profiles by binning member stars in constant width
bins in \(\log R_{\rm ell}\) of 0.05 dex. We ignore bins interior or
exterior to the first empty bin in either direction. We use symmetric
Poisson uncertainties as ``error bars'' in the density estimate at each
bin.

Fig.~\ref{fig:scl_observed_profiles} show the derived density profiles
for Sculptor, Ursa Minor, and Fornax. We calculate density profiles for
different selections of stars from above: \emph{all} stars, CMD+PM only,
and the fiducial sample. In each case, all samples are the same towards
the inner regions of the satellite. However, the density profile
\emph{quality stars} plateaus at the total background in the field at
radii of 30-60 arcminutes, depending on the galaxy. By restricting
stellar membership with CMD + PM, the background is reduced by 1-2 dex.
The CMD+PM background plateau represents the density of background stars
which could be mistaken as members because of their coincident colours
and proper motions. Finally, the background (BG)-subtracted profile
results from subtracting the apparent background in the \emph{all}
profile.

\textbf{Describe R\_break, etc. in Fig.2.4} Note that, in all three
galaxies, the data does not extend out to the break radius.

While the density profiles of all samples agrees in the inner regions,
they begin to deviate outside \(6 R_h\). Outside this radius, the
fiducial profile extends 1--2 magnitudes below the CMD+PM selection
background. Because the fiducial profile in this region depends
critically on the spatial likelihood, its shape is vulnerable to the
assumed density profile (an exponential in J+24). We explore this in
more detail in Appendix {[}@-sec:density\_extra{]}.

Nearto UMi, there is a small, likely unassociated, ultrafaint star
cluster, Muñoz 1 \citep{munoz+2012}. The cluster is at a relative
position of \((\xi, \eta) \approx(-42, -15)\) arcminutes, corresponding
to an elliptical radius of 36 arcminutes. This cluster does not have a
bright RGB, so has few stars brighter than a magnitude \(G=22\). The
cluster has little effect on the elliptically-averaged density profile.
\textbf{Mark Muñoz 1}

Fig.~\ref{fig:classical_dwarfs_densities} compares the fiducial density
profiles of Sculptor, Ursa Minor, Fornax, and other classical dwarf
galaxies. Of the classicals, we exclude Antlia II due to the extremely
high background, and Sagittarius which was not included in J+24. The
density profiles are scaled to match at the half-light radius from
\citet{munoz+2018}. All of the classical dwarfs appear to be well
described by an exponential profile in the inner regions. In the outer
regions, however, Sculptor and Ursa Minor deviate, and show a clear
outer excess over an exponential law (solid black line). These galaxies
are better fit by a Plummer law (dashed black line). The deviation from
an exponential grows outwards, and at \(\sim 8 R_h\), may reach 2 orders
of magnitude. This excess is remarkable because the spatial likelihood
used to derive the fiducial sample probabilities is the same (a single
component exponential) in all cases. The remainder of this thesis will
be devoted to assessing whether the outer excess shown by Scl and UMi
are due to Galactic tides.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_fnx_density_methods.pdf}}
\caption[Sculptor density profiles]{The density profile of Sculptor,
Ursa Minor, and Fornax for different selection criteria, ploted as log
surface dencity versus log elliptical radius. \emph{CMD+PM} select stars
more likely to be members according to CMD and PM only, \emph{all}
selects any high quality star, and \emph{BG subtracted} is the
background-subtracted density derived from \emph{all} stars. We mark the
half-light radius (vertical dashed line) and the break radius (black
arrow, Eq.~\ref{eq:r_break}). \textbf{rename
bg-subtracted}}\label{fig:scl_observed_profiles}
\end{figure}

@@
